---
title: "Statistics From Scratch (Theory + R + ML)"
author: "Venkateswara Rao Jannegorla"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
---

# Welcome

This book builds statistics from foundations to machine learning connections.

It is designed for:

- Beginners
- Data science students
- Business analytics learners
- ML engineers who want strong statistical foundations

# How to Use This Book

1. Read intuition carefully  
2. Run the R code  
3. Think about business application  
4. Connect to ML concepts  

# Roadmap

1. Introduction  
2. Data & Collection  
3. Descriptive Statistics  
4. Probability  
5. Inference  
6. Regression  
7. Time Series  
8. Advanced Topics  

<!--chapter:end:index.Rmd-->

---
title: "Chapter 1: Introduction to Statistics"
author: "Your Name"
date: "`r Sys.Date()`"
output: bookdown::html_document2
---


---
title: "Chapter 1: Introduction to Statistics"
author: "Your Name"
date: "`r Sys.Date()`"
output: bookdown::html_document2
---

# 1. Why Does Uncertainty Exist?

Every day we make decisions without knowing the future.

- Will tomorrow’s sales increase?
- Will it rain?
- Will a patient recover?
- Will a machine fail?

The world is not deterministic.  
It contains **variation**, **uncertainty**, and **randomness**.

Even if we repeat the same action twice, results can differ.

For example:

- Two students study equal hours but score differently.
- A marketing campaign works in one city but fails in another.
- A model performs well on training data but poorly on new data.

This variability is the starting point of statistics.

Statistics exists because the world is not perfectly predictable.

---

# 2. What Is Data?

Data is recorded information about the world.

It can represent:

- Sales numbers
- Heights of students
- Temperature readings
- Website clicks
- Medical test results

But raw data alone has no meaning.

Consider this vector:

```{r}
sales <- c(120, 135, 150, 98, 175, 142)
sales
```

These are just numbers.

Without context, they tell us nothing.

Statistics gives structure and meaning to data.

---

# 3. Why Raw Data Is Not Enough

Imagine you are a business manager.

You receive 10,000 rows of customer transaction data.

What do you want to know?

- Average purchase amount
- Most common product category
- Seasonal trends
- Whether marketing increased revenue

Raw data does not answer these.

We need methods to:

1. Summarize
2. Analyze
3. Interpret
4. Draw conclusions

That system of methods is statistics.

---

# 4. What Is Statistics?

Statistics is the science of learning from data.

More formally:

It is the discipline concerned with collecting, organizing, analyzing, interpreting, and presenting data to make decisions under uncertainty.

Statistics has two main branches:

1. Descriptive Statistics
2. Inferential Statistics

---

# 5. Descriptive vs Inferential Statistics

## 5.1 Descriptive Statistics

Descriptive statistics summarizes data.

Examples:

- Mean (average)
- Median
- Standard deviation
- Charts and graphs

Example in R:

```{r}
mean(sales)
sd(sales)
```

These numbers describe only the data we have.

They do not generalize beyond it.

---

## 5.2 Inferential Statistics

Inferential statistics goes further.

It asks:

"If this sample looks like this, what can we say about the whole population?"

Example:

We survey 200 customers.
Can we say something about all 50,000 customers?

Inferential statistics allows us to:

- Estimate unknown quantities
- Test hypotheses
- Measure uncertainty
- Build predictive models

This is where probability becomes important.

---

# 6. Population vs Sample

Suppose an e-commerce company has:

- 1,000,000 customers (population)

We survey:

- 2,000 customers (sample)

Why not ask all 1,000,000?

Because:

- It is expensive
- It is slow
- It may be impractical

Statistics allows us to use samples to understand populations.

This idea is the foundation of modern data science.

---

# 7. Where Statistics Is Used in Business

Statistics is everywhere in business:

### Marketing
- A/B testing
- Conversion rate optimization

### Finance
- Risk modeling
- Portfolio optimization

### Operations
- Demand forecasting
- Inventory control

### Healthcare
- Drug effectiveness testing
- Survival analysis

### Technology
- Recommendation systems
- Fraud detection

Without statistics, decisions become guesses.

---

# 8. Statistics vs Probability

These two are closely related but different.

Probability:
- Starts with a known model
- Predicts outcomes

Statistics:
- Starts with data
- Infers the model

Example:

Probability:
"What is the chance of getting heads?"

Statistics:
"Based on 100 flips, is this coin biased?"

Probability moves forward.
Statistics moves backward.

Machine learning combines both.

---

# 9. Case Study: Online Retail Company

Imagine an online store.

They observe:

```{r}
daily_sales <- c(120, 150, 130, 160, 200, 170, 180)
mean(daily_sales)
```

The average daily sales is around 158.

But management asks:

- Is this stable?
- Is there growth?
- Is advertising working?
- Can we predict next month?

Now we need:

- Variability analysis
- Trend analysis
- Hypothesis testing
- Regression modeling

Statistics transforms data into strategy.

---

# 10. Statistics in Machine Learning

Machine learning is built on statistical ideas.

Examples:

- Linear regression = statistical regression model
- Logistic regression = probability model
- Naive Bayes = conditional probability
- Gaussian distributions = normal distribution theory
- Overfitting = variance concept
- Bias-variance tradeoff = statistical error decomposition

Even neural networks rely on:

- Optimization
- Loss functions
- Probability distributions

If statistics is weak,
machine learning becomes memorization without understanding.

Strong statisticians become strong ML engineers.

---

# 11. The Big Picture

Statistics answers four fundamental questions:

1. What happened?
2. Why did it happen?
3. What will happen next?
4. How confident are we?

It provides tools to reason under uncertainty.

It replaces guesswork with structured thinking.

---

# 12. What You Will Learn in This Book

In the next chapters, we will explore:

- Data types and measurement scales
- Sampling techniques
- Descriptive measures
- Probability theory
- Distributions
- Estimation and confidence intervals
- Hypothesis testing
- Regression and correlation
- Time series analysis
- Non-parametric methods
- Advanced statistical concepts used in AI

Each concept will include:

- Intuition
- Real-world application
- Mathematical explanation
- R implementation
- ML connection

---

# Final Thought

Statistics is not about numbers.

It is about understanding uncertainty.

Once you understand uncertainty,
you understand risk,
and once you understand risk,
you make better decisions.

This is why statistics matters.

<!--chapter:end:book/01_introduction.Rmd-->

