# Chapter 2: Data and Data Collection

Data is the raw material of statistics.

Without good data, even the best statistical methods fail.

---

## What Is Data?

Data is recorded information about the world.

Examples:

- Daily sales numbers  
- Temperature readings  
- Customer reviews  
- Blood pressure measurements  
- Website clicks  

Understanding what data represents is the first step toward correct analysis.

---

## Types of Data

There are two broad categories of data.

### Qualitative (Categorical) Data

This describes qualities or categories.

Examples:

- Gender  
- Product category  
- Blood type  
- Customer satisfaction level  

Example in R:

```{r}
category <- c("Electronics", "Clothing", "Electronics", "Groceries")
table(category)
```

This data describes groups, not numerical measurements.

---

### Quantitative (Numerical) Data

This represents measurable quantities.

Examples:

- Sales amount  
- Height  
- Weight  
- Revenue  

Example:

```{r}
sales <- c(120, 135, 150, 98, 175)
mean(sales)
```

Quantitative data allows mathematical operations.

---

## Levels of Measurement

The level of measurement determines what type of analysis is valid.

### Nominal Scale

Categories without order.

Example:
- Colors  
- Blood type  

You cannot compute an average of colors.

---

### Ordinal Scale

Categories with order but no fixed difference.

Example:
- Satisfaction: Low, Medium, High  
- Education level  

We know the order, but differences are not measurable.

---

### Interval Scale

Ordered and measurable but no true zero.

Example:
- Temperature in Celsius  

0°C does not mean “no temperature.”

---

### Ratio Scale

Ordered, measurable, true zero.

Example:
- Income  
- Height  
- Sales revenue  

Most statistical methods require ratio scale data.

---

## Population vs Sample

### Population

Entire group of interest.

Example:
All customers of an e-commerce company.

### Sample

Subset of population used for analysis.

Example:
Surveying 2,000 customers from 1 million.

Why use samples?

- Cheaper  
- Faster  
- Practical  

Statistics allows us to generalize from sample to population.

---

## Census vs Sampling

### Census

Collect data from entire population.

Example:
National census.

### Sampling

Collect data from a subset.

Most businesses use sampling.

Sampling is powerful — but dangerous if done incorrectly.

---

## Sampling Techniques

### Simple Random Sampling

Every individual has equal chance.

Example simulation:

```{r}
set.seed(123)
population <- 1:10000
sample_data <- sample(population, 100)
head(sample_data)
```

---

### Stratified Sampling

Divide population into groups, then sample from each group.

Example:
- Male / Female  
- Region  
- Age group  

Used to ensure representation.

---

### Systematic Sampling

Select every k-th element.

Example:
Every 10th customer.

---

### Cluster Sampling

Select entire groups randomly.

Example:
Choose 5 stores and survey all customers there.

---

## Sampling Bias

Bias occurs when the sample does not represent the population.

Example:

Survey only online users → ignores offline customers.

In machine learning:

If training data is biased, the model becomes biased.

Real-world example:
Facial recognition systems failing due to non-diverse training data.

Bias leads to systematic error.

---

## Types of Errors

### Sampling Error

Difference between sample statistic and population parameter.

Unavoidable but measurable.

---

### Non-Sampling Error

- Data entry mistakes  
- Poor survey design  
- Measurement error  

Often more dangerous than sampling error.

---

## Business Case Study

Imagine an online retailer.

They survey only customers who purchased in the last week.

Is this sample representative?

No.

Inactive customers are ignored.

Marketing decisions based on this data may be flawed.

---

## ML Dataset Bias Example

Suppose you train a loan approval model using:

- Mostly urban applicants  
- Few rural applicants  

The model learns patterns biased toward urban behavior.

Result:
Unfair decisions.

Statistics teaches us:

Data collection quality determines model fairness.

---

## Why This Chapter Matters

Most analytical errors do not come from complex math.

They come from:

- Poor sampling  
- Biased data  
- Misunderstood measurement scales  

If data is wrong, analysis will be wrong.

Garbage in → garbage out.

---

## Summary

In this chapter we learned:

- Types of data  
- Measurement scales  
- Population vs sample  
- Sampling techniques  
- Bias and errors  
- Business implications  
- Machine learning fairness connection  

Next: Descriptive statistics.