---
title: "Chapter 2: Data and Data Collection"
---

# 1. What Is Data?

Data is recorded information about the world.

Examples:

- Daily sales numbers
- Temperature readings
- Customer reviews
- Blood pressure measurements
- Website clicks

But not all data is the same.

Understanding data types is the first step toward correct analysis.

---

# 2. Types of Data

There are two broad categories of data.

## 2.1 Qualitative (Categorical) Data

This describes qualities or categories.

Examples:

- Gender
- Product category
- Blood type
- Customer satisfaction level

Example in R:

```{r}
category <- c("Electronics", "Clothing", "Electronics", "Groceries")
table(category)
```

This data describes groups, not numerical measurements.

---

## 2.2 Quantitative (Numerical) Data

This represents measurable quantities.

Examples:

- Sales amount
- Height
- Weight
- Revenue

Example:

```{r}
sales <- c(120, 135, 150, 98, 175)
mean(sales)
```

Quantitative data allows mathematical operations.

---

# 3. Levels of Measurement

Understanding measurement scale determines what analysis is valid.

## 3.1 Nominal Scale

Categories without order.

Example:
- Colors
- Blood type

You cannot compute average of colors.

---

## 3.2 Ordinal Scale

Categories with order but no fixed difference.

Example:
- Satisfaction: Low, Medium, High
- Education level

Difference between levels is not measurable.

---

## 3.3 Interval Scale

Ordered and measurable but no true zero.

Example:
- Temperature in Celsius

0°C does not mean "no temperature".

---

## 3.4 Ratio Scale

Ordered, measurable, true zero.

Example:
- Income
- Height
- Sales revenue

Most statistical methods require ratio scale data.

---

# 4. Population vs Sample

## 4.1 Population

Entire group of interest.

Example:
All customers of an e-commerce company.

## 4.2 Sample

Subset of population used for analysis.

Example:
Surveying 2,000 customers from 1 million.

Why sample?

- Cheaper
- Faster
- Practical

Statistics allows us to generalize from sample to population.

---

# 5. Census vs Sampling

## Census

Collect data from entire population.

Example:
National census.

## Sampling

Collect data from subset.

Most businesses use sampling.

Sampling is powerful — but dangerous if done incorrectly.

---

# 6. Sampling Techniques

## 6.1 Simple Random Sampling

Every individual has equal chance.

Example simulation:

```{r}
set.seed(123)
population <- 1:10000
sample_data <- sample(population, 100)
head(sample_data)
```

---

## 6.2 Stratified Sampling

Divide population into groups, then sample from each.

Example:
- Male / Female
- Region
- Age group

Used to ensure representation.

---

## 6.3 Systematic Sampling

Select every k-th element.

Example:
Every 10th customer.

---

## 6.4 Cluster Sampling

Select entire groups randomly.

Example:
Choose 5 stores and survey all customers there.

---

# 7. Sampling Bias

Bias occurs when sample does not represent population.

Example:

Survey only online users → ignores offline customers.

In ML:

Training data biased → model biased.

Real-world example:
Facial recognition systems failing due to non-diverse training data.

Bias leads to systematic error.

---

# 8. Types of Errors

## 8.1 Sampling Error

Difference between sample statistic and population parameter.

Unavoidable but measurable.

## 8.2 Non-Sampling Error

- Data entry mistakes
- Poor survey design
- Measurement error

Often more dangerous than sampling error.

---

# 9. Business Case Study

Imagine an online retailer.

They survey only customers who purchased in the last week.

Question:
Is this sample representative?

No.

Inactive customers are ignored.

Conclusion:
Marketing decisions based on this data may be flawed.

---

# 10. ML Dataset Bias Example

Suppose you train a loan approval model using:

- Mostly urban applicants
- Few rural applicants

Model learns patterns biased toward urban behavior.

Result:
Unfair decisions.

Statistics teaches us:

Data collection quality determines model fairness.

---

# 11. Why This Chapter Matters

Most analytical errors do NOT come from complex math.

They come from:

- Poor sampling
- Biased data
- Misunderstood measurement scales

If data is wrong,
analysis will be wrong.

Garbage in → garbage out.

---

# 12. Summary

In this chapter we learned:

- Types of data
- Measurement scales
- Population vs sample
- Sampling techniques
- Bias and errors
- Business implications
- ML fairness connection

Next:
We move to descriptive statistics.